---
title: "Create a distance raster from spatial points"
author: "Ricardo Ochoa, Eduardo Denicia"
date: "8/28/2018"
output: html_document
---

This tutorial will guide you in a step-by-step process to create create a distance raster from spatial points. This tutorial considers that you are already familiar with the following concepts. If you are not, please visit the links.

- [R](https://www.r-project.org/)
- [RStudio](https://www.rstudio.com)
- [Raster](https://en.wikipedia.org/wiki/Raster_graphics)

## Introduction 
Urban studies commonly depend on assessing how city services are distributed in the city. However, technical teams commonly struggle to generate, maintain or retrive detailed spatial information. The following R code explains how to create a **raster** (GeoTif) in which each pixel includes the minimum distance (crow-flight) to the closest hospital. In this tutorial we will use data from hospitals in Mexico City and the output from the [Create a raster with built-up area from GHSL data](https://github.com/ricardoochoa/ST_preprocessing/tree/master/build_up_area) tutorial.

## 1. Load required libraries
In this tutorial we will use the following libraries:
[raster](https://cran.r-project.org/web/packages/raster/raster.pdf): Title Geographic Data Analysis and Modeling Version 2.6-7, November 12, 2017
[rgdal](https://cran.r-project.org/web/packages/rgdal/rgdal.pdf): Bindings for the 'Geospatial' Data Abstraction, Version 1.3-4, August 3, 2018
[sp](https://cran.r-project.org/web/packages/sp/sp.pdf): Classes and Methods for Spatial Data, Version 1.3-1, June 5, 2018
[leaflet](https://cran.r-project.org/web/packages/leaflet/leaflet.pdf): Create Interactive Web Maps with the JavaScript 'Leaflet', Version 2.0.2, August 27, 2018

Run the following lines to load the libraries. 

```{r 2 libraries, include=TRUE, echo=TRUE, message=FALSE}
# install.packages('raster')    # Run only once
# install.packages('rgdal')     # Run only once
# install.packages('sp')        # Run only once
# install.packages('leaflet')   # Run only once
library(raster)
library(rgdal)
library(sp)
library(leaflet)
```

## 2. Prepare data
Please download the data from our GitHub repository: [distance_raster](https://github.com/ricardoochoa/ST_preprocessing/tree/master/distance_raster). Notice that you will need the contents of the _input file. As you will notice inputs include a TIF file (raster) with the built-up are in Mexico City Metro Area and a CSV file, with the location of all public hospitals in Mexico City. 

First of all, we will import and explore the data. 

```{r rasterVis, echo=FALSE}

hospitals <- read.csv("_input/hospitals.csv")
built_up  <- raster("_input/built_up_mexico_city.tif")

leaflet(hospitals) %>% addTiles() %>%
  addMarkers(lng = ~lng, lat = ~lat) %>%
  addRasterImage(built_up, opacity = 0.5)

```



## 3. Estimate the distance
```{r estimate distance}
ghsl_crs <- '+proj=moll +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs'

points <- SpatialPointsDataFrame(coords = hospitals[c("lng", "lat")], data = hospitals,
                               proj4string = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))

points <- spTransform(x = points, CRSobj = ghsl_crs)

distance_raster <- distanceFromPoints(object = built_up, xy = points)
```

## 4. Plot
```{r plot raster}
leaflet() %>% addTiles() %>%
  addRasterImage(distance_raster, opacity = 0.5)

```


## 5. Save your raster file

Save your raster file in your preffered directory.

```{r write raster}
writeRaster(distance_raster, filename = "_output/distance_raster.tif", overwrite = TRUE)
```

That's it!

