---
title: "Create a distance raster from spatial points in R"
author: "Ricardo Ochoa, Eduardo Denicia"
date: "8/28/2018"
output: html_document
---

This tutorial will guide you in a step-by-step process to create create a distance raster from spatial points. This tutorial considers that you are already familiar with the following concepts. If you are not, please visit the links.

- [R](https://www.r-project.org/)
- [RStudio](https://www.rstudio.com)
- [Raster](https://en.wikipedia.org/wiki/Raster_graphics)

## Introduction 
Urban studies commonly depend on assessing how city services are distributed in the city. However, technical teams commonly struggle to generate, maintain or retrive detailed spatial information. The following R code explains how to create a **raster** (GeoTif) in which each pixel includes the minimum distance (crow-flight) to the closest hospital. In this tutorial we will use data from hospitals in Mexico City and the output from the [Mask and crop raster from shapefile](https://github.com/ricardoochoa/ST_preprocessing/tree/master/mask_and_crop_raster_from_shapefile) tutorial.

## 1. Load required libraries
In this tutorial we will use the following libraries:
[raster](https://cran.r-project.org/web/packages/raster/raster.pdf): Title Geographic Data Analysis and Modeling Version 2.6-7, November 12, 2017
[rgdal](https://cran.r-project.org/web/packages/rgdal/rgdal.pdf): Bindings for the 'Geospatial' Data Abstraction, Version 1.3-4, August 3, 2018
[sp](https://cran.r-project.org/web/packages/sp/sp.pdf): Classes and Methods for Spatial Data, Version 1.3-1, June 5, 2018
[leaflet](https://cran.r-project.org/web/packages/leaflet/leaflet.pdf): Create Interactive Web Maps with the JavaScript 'Leaflet', Version 2.0.2, August 27, 2018

Run the following lines to load the libraries. 

```{r 2 libraries, include=TRUE, echo=TRUE, message=FALSE}
# install.packages('raster')    # Run only once
# install.packages('rgdal')     # Run only once
# install.packages('sp')        # Run only once
# install.packages('leaflet')   # Run only once
library(raster)
library(rgdal)
library(sp)
library(leaflet)
```

## 2. Prepare data
If you want to replicate this tutorial, please download the data from our GitHub repository: [distance_raster](https://github.com/ricardoochoa/ST_preprocessing/tree/master/distance_raster). Notice that you will need the contents of the _input file. As you will notice inputs include a TIF file (raster) with the built-up are in Mexico City, CDMX and a CSV file, with the location of all [public hospitals in CDMX](www.datosabiertos.cdmx.gob.mx:5000/it/dataset/hospitales-y-centros-de-salud). 

First of all, we will import and explore the data.

```{r rasterVis}
built_up  <- raster("_input/cdmx_builtup.tif")
plot(built_up)

hospitals <- read.csv("_input/hospitals.csv")
plot(hospitals[c("lng", "lat")])

```

Please note from both plots that **hospitals** and **built-up** have different CRS. 

## 3. Estimate the distance
The distance estimation requires only a couple of lines. However, you might need to transform your data to match CRS between layers. In our example we will transform all hospital points to the built-up raster's CRS. 

```{r fix crs}
ghsl_crs <- '+proj=moll +lon_0=0 +x_0=0 +y_0=0 +ellps=WGS84 +datum=WGS84 +units=m +no_defs'

# Read the hospitals data 
points <- SpatialPointsDataFrame(coords = hospitals[c("lng", "lat")], data = hospitals,
                               proj4string = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))

# Transform the CRS
points <- spTransform(x = points, CRSobj = ghsl_crs)

```
Once that we have the same CRS in both sources, we can estimate the distances with the following code. Please note that the **distanceFromPoints** function will estimate the shortest distance (in meters) to the nearest hospital. 

```{r estimate distance}
# Distance from points to each raster cell
distances <- distanceFromPoints(object = built_up, xy = points)
plot(distances)

# Lastly, we will "mask" the raster to keep only those cells inside CDMX
distance_raster <- mask(x = distances, mask = built_up)
plot(distance_raster)
```

## 4. Plot
Lastly, we will plot the final result using leaflet. 

```{r plot raster, warning = FALSE}

pal <- colorNumeric(c("#556270", "#4ECDC4", "#C7F464", "#FF6B6B", "#C44D58"),
                    values(distance_raster),
                    na.color = "transparent")

leaflet() %>%
  addProviderTiles(providers$OpenStreetMap.BlackAndWhite) %>%
  addRasterImage(distance_raster, colors = pal, opacity = 0.5) %>%
  addLegend(pal = pal, values = values(distance_raster),
    title = "Distance in meters")

```

## 5. Save your raster file

Save your raster file in your preffered directory.

```{r write raster}
writeRaster(distance_raster, filename = "_output/distance_raster.tif", overwrite = TRUE)
```

That's it!

